---
title: 'Replicating "What is the best measure of labor market tightness?"'
author: "Itamar Caspi"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    highlight: haddock
    keep_md: no
    theme: journal
    toc: yes
    toc_depth: 4
    toc_float: yes
abstract: |
  This notebook includes replication of some of the regression results that appear in ["What is the best measure of labor market tightness?"](https://www.piie.com/blogs/realtime-economic-issues-watch/what-best-measure-labor-market-tightness) by [Jason Furman](https://www.piie.com/experts/senior-research-staff/jason-furman) and [Willson Powell III](https://www.piie.com/experts/former-research-staff/wilson-powell-iii). 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


## Load packages
```{r}
library(tidyverse)   # for data wrangling and plots
library(lubridate)   # for dealing with dates
library(fredr)       # a FRED api
library(broom)       # tidying regression output
library(hrbrthemes)  # fancy plots
library(gghighlight) # for labeling series in plots
library(gt)          # fancy tables
```


## Download data from [FRED](https://fred.stlouisfed.org/)

Note: To run this code, you'll first need to get a [FRED API key](https://fred.stlouisfed.org/docs/api/api_key.html) and then set it using the `fredr_set_key()` function.


Download data
```{r}
params <- list(
  series_id = c(
    "PAYEMS", "UNRATE", "LNS12300060", "JTSQUL", "JTSJOL", "UNEMPLOY", "CPILFESL", "PCEPILFE"
    ),
  frequency = c("q", "q", "q", "q", "q", "q", "q", "q"),
  units = c("lin", "lin", "lin", "lin", "lin", "lin", "pc1", "pc1")
)

df_raw <- pmap_dfr(
  .l = params,
  .f = ~ fredr(series_id = ..1, frequency = ..2, units = ..3)
)
```

## Preprocess data

Change to "wide" format
```{r}
df_wide <- 
  df_raw %>% 
  select(date, series_id, value) %>% 
  pivot_wider(names_from = "series_id", values_from = "value")

head(df_wide)
```

Calculate and rename variables
```{r}
df <- 
  df_wide %>%
  mutate(
    ur       = UNRATE,
    panr     = 100 - LNS12300060,
    upjo     = UNEMPLOY / JTSJOL,
    qr       = 100 - 100 * JTSQUL/PAYEMS,
    core_cpi = CPILFESL,
    core_pce = PCEPILFE
  ) %>% 
  select(date, ur, panr, upjo, qr, core_cpi, core_pce) %>% 
  filter(between(date, ymd("2001-01-01"), ymd("2021-07-31")))

head(df)
```

Transformer variables to Z-scores based on a mean and standard deviation from 2001 to 2018.

First, get mean and standard deviation
```{r}
params <- 
  df %>%
  filter(date <= ymd("2018-12-31")) %>% 
  summarise(across(ur:core_pce, list(mean = mean, sd = sd)))

params
```


```{r}
# generate z-scores
df_zscore <- 
  df %>% 
  mutate(
    ur       = (ur - params$ur_mean) / params$ur_sd,
    panr     = (panr - params$panr_mean) / params$panr_sd,
    upjo     = (upjo - params$upjo_mean) / params$upjo_sd,
    qr       = (qr - params$qr_mean) / params$qr_sd,
    core_cpi = (core_cpi - params$core_cpi_mean) / params$core_cpi_sd,
    core_pce = (core_pce - params$core_pce_mean) / params$core_pce_sd
  )

head(df_zscore)
```


## Plot time series

Labor market indicators
```{r}
df_zscore %>% 
  select(-starts_with("core")) %>% 
  pivot_longer(-date, names_to = "name", values_to = "value") %>% 
  ggplot(aes(date, value, color = name)) + 
  geom_line(size = 1) +
  scale_color_viridis_d() +
  gghighlight(max(value) > -999) +
  geom_hline(aes(yintercept = 0)) +
  theme_ipsum() +
  labs(
    subtitle = "Quarterly measures of labor market tightness",
    x = "",
    y = "Z-score"
  )
```

Inflation indicators
```{r}
df_zscore %>% 
  select(date, starts_with("core")) %>% 
  pivot_longer(-date, names_to = "name", values_to = "value") %>% 
  ggplot(aes(date, value, color = name)) + 
  geom_line(size = 1) +
  scale_color_viridis_d() +
  gghighlight(max(value) > -999) +
  geom_hline(aes(yintercept = 0)) +
  theme_ipsum() +
  labs(
    subtitle = "Percent Change in measurments of prices over past 4 quarters",
    x = "",
    y = "Z-score"
  )
```


## Regression analysis

We now estimate a traditional and an autoregressive Phillips curve that relates current measures of slack to changes in prices  over the next year:

$$
\text { Inflation growth }_{\mathrm{t} \text { to } \mathrm{t}+4 \mathrm{q}}=\beta_{0}+\beta_{1} * \text { Slack }_{t}+\varepsilon_t
$$
and
$$
\text { Inflation growth }_{\mathrm{t} \text { to } \mathrm{t}+4 \mathrm{q}}=\beta_{0}+\beta_{1} * \text { Slack }_{t}+\text { Inflation growth }_{\mathrm{t}-4\mathrm{q}\text { to } \mathrm{t}}\varepsilon_t
$$

First, nest the data to make easier to "map" over the different models
```{r}
df_nest <- 
  df_zscore %>% 
  filter(date <= ymd("2019-12-31")) %>% 
  pivot_longer(-c(date, core_cpi, core_pce) ,names_to = "name", values_to = "value") %>% 
  group_by(name) %>% 
  nest()

head(df_nest)
```

Next, run the regressions and store the results
```{r}
df_results <- 
  df_nest %>% 
  mutate(
    model_cpi    = map(data, ~ lm(core_cpi ~ lag(value, 4), data = .)),
    model_pce    = map(data, ~ lm(core_pce ~ lag(value, 4), data = .)),
    model_cpi_ar = map(data, ~ lm(core_cpi ~ lag(value, 4) + lag(core_cpi, 4), data = .)),
    model_pce_ar = map(data, ~ lm(core_pce ~ lag(value, 4) + lag(core_pce, 4), data = .)),
    stats_cpi    = map(model_cpi, ~ glance(.)),
    stats_pce    = map(model_pce, ~ glance(.)),
    stats_cpi_ar = map(model_cpi_ar, ~ glance(.)),
    stats_pce_ar = map(model_pce_ar, ~ glance(.))
  ) %>% 
  unnest(stats_cpi, stats_pce, stats_cpi_ar, stats_pce_ar) %>%
  select(name, adj.r.squared, adj.r.squared1, adj.r.squared2, adj.r.squared3)
```


## Results

Print results with the help of the [`gt`](https://gt.rstudio.com/) package
```{r}
df_results %>%
  ungroup() %>% 
  gt() %>% 
  tab_header(
    title = "Adjusted R squared from Phillips cureve regressions"
  ) %>% 
  cols_label(
    name           = "Dependent var.",
    adj.r.squared  = "Core CPI",
    adj.r.squared1 = "Core PCE",
    adj.r.squared2 = "Core CPI",
    adj.r.squared3 = "Core PCE"
  ) %>% 
  tab_spanner(
    label = "Traditional",
    columns = c(adj.r.squared, adj.r.squared1)
  ) %>%
  tab_spanner(
    label = "Autoregressive",
    columns = c(adj.r.squared2, adj.r.squared3)
  ) %>% 
  fmt_number(
    columns = starts_with("adj."),
    decimals = 2
  )
```

For more results and further details, please refer to the [full article](https://www.piie.com/blogs/realtime-economic-issues-watch/what-best-measure-labor-market-tightness).